---
# YAML is the control panel. It does not run R.
# It tells Quarto the title, the output format, and default chunk rules.
title: "Fuel Efficiency in Motor Vehicles: An Exploratory Report"
subtitle: "A Sample Report Demonstrating Quarto Features"
author: "Student Name"
date: today                          # filled in when you Render
format:
  pdf:                               # change to html if PDF / LaTeX fails
    toc: true                        # table of contents from # / ## headings
    number-sections: true            # 1, 1.1, 2, ... on headings
    number-depth: 3                  # number down to ### as well
    fig-width: 6                     # default figure width (inches)
    fig-height: 4
    colorlinks: true                 # blue links in the PDF
execute:
  echo: true                         # show code unless a chunk overrides this
  warning: false                     # hide warnings (package startup, etc.)
  message: false                     # hide messages (readr column types, etc.)
---

::: {.callout-note}
This is a sample, not an assignment. Comments that start with `#` explain
what Quarto is doing. You can delete them when you write your own report.
Render the file from an RStudio project. If PDF fails, change `format: pdf`
to `format: html` and click Render again.
:::

```{r setup}
#| include: false
# include: false → R still runs this chunk, but the reader sees neither
# the code nor the output. Use this for libraries and data prep.

# Load packages. Do this once per document, near the top.
# Do not put install.packages() here. Install once in the console.
library(tidyverse)
library(knitr)
library(broom) # turns model output into a table; install once if needed

# mtcars ships with R, so this report needs no external CSV.
data(mtcars)
# rownames_to_column() makes the car names an ordinary column.
mtcars <- mtcars |> rownames_to_column(var = "model")
```

# Introduction

Fuel efficiency is one of the most important factors car buyers consider.
In this report, we explore the relationship between a vehicle's weight and
its fuel efficiency using the classic `mtcars` dataset, which contains
specifications for `r nrow(mtcars)` car models from a 1974 *Motor Trend*
magazine issue.

Notice that the number of cars above (`r nrow(mtcars)`) was **not typed
manually** — it was inserted with inline R code
(`` `r knitr::inline_expr("nrow(mtcars)")` ``). This is one of the most
useful habits to build: numbers that come from your data should be computed,
not hard-coded, so the report updates automatically if the data changes.

# Data Overview

## Variable Description

@tbl-vars describes the variables used in this report.
<!-- @tbl-vars is a cross-reference. It becomes a clickable "Table 1" because
     the next chunk has label: tbl-vars and tbl-cap. -->

```{r}
#| label: tbl-vars
#| tbl-cap: "Description of key variables in the mtcars dataset"
# label + tbl-cap together create a numbered table you can cite with @tbl-...

var_table <- tibble(
  Variable = c("mpg", "wt", "hp", "cyl"),
  Description = c(
    "Miles per gallon (fuel efficiency)",
    "Weight (in 1000 lbs)",
    "Gross horsepower",
    "Number of cylinders"
  )
)

kable(var_table) # knitr::kable() prints a clean table in HTML or PDF
```

## A Quick Look at the Data

Here are the first few rows of the raw data, shown in @tbl-head.

```{r}
#| label: tbl-head
#| tbl-cap: "First six cars in the dataset"

kable(head(mtcars[, c("model", "mpg", "wt", "hp", "cyl")]))
```

# Exploratory Analysis

## Distribution of Fuel Efficiency

The average fuel efficiency across all cars is
`r round(mean(mtcars$mpg), 1)` miles per gallon, with a standard deviation
of `r round(sd(mtcars$mpg), 1)`.
The two numbers in that sentence are inline R, not typed by hand.
@fig-hist shows the full distribution.

```{r}
#| label: fig-hist
#| fig-cap: "Distribution of miles per gallon across all cars"
# label + fig-cap create a numbered figure you can cite with @fig-...

ggplot(mtcars, aes(x = mpg)) +
  geom_histogram(binwidth = 3, fill = "steelblue", color = "white") +
  labs(x = "Miles per gallon", y = "Count") +
  theme_minimal()
```

## Weight, Horsepower, and Efficiency

To compare two relationships at once, we place two plots side by side using
`layout-ncol`, shown in @fig-side.

```{r}
#| label: fig-side
#| fig-cap: "Weight and horsepower versus fuel efficiency"
#| layout-ncol: 2
#| fig-subcap:
#|   - "Weight vs. MPG"
#|   - "Horsepower vs. MPG"
# Two ggplot() calls in one chunk + layout-ncol: 2 → two panels.
# fig-subcap labels each panel. The chunk still has one overall @fig-side.

ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point(color = "darkred") +
  labs(x = "Weight (1000 lbs)", y = "MPG") +
  theme_minimal()

ggplot(mtcars, aes(x = hp, y = mpg)) +
  geom_point(color = "darkgreen") +
  labs(x = "Horsepower", y = "MPG") +
  theme_minimal()
```

Both panels suggest a **negative relationship**: heavier, more powerful
cars tend to be less fuel efficient.

# Statistical Model

## Model Specification

We fit a simple linear regression of fuel efficiency on weight:

$$
\text{mpg}_i = \beta_0 + \beta_1 \cdot \text{wt}_i + \varepsilon_i
$$

<!-- Dollar signs mark LaTeX math. One $ is inline; two $$ is a centered equation. -->

where $\beta_0$ is the intercept, $\beta_1$ is the slope, and
$\varepsilon_i$ is the error term for car $i$.

## Model Results

```{r}
#| label: tbl-model
#| tbl-cap: "Linear regression results: MPG explained by weight"

model <- lm(mpg ~ wt, data = mtcars)
# broom::tidy() puts the coefficients in a data frame kable() can print.
kable(broom::tidy(model), digits = 3)
```

The fitted model can be written as:

$$
\widehat{\text{mpg}} = `r round(coef(model)[1], 2)` \; `r round(coef(model)[2], 2)` \times \text{wt}
$$

<!-- Inline R can sit inside a math equation. coef(model) comes from the
     chunk above; Render runs the file from top to bottom. -->

This means that for every additional 1,000 lbs of weight, fuel efficiency
drops by about `r abs(round(coef(model)[2], 2))` miles per gallon, holding
all else constant.

## Model Fit

```{r}
#| label: fig-fit
#| fig-cap: "Fitted regression line over the observed data"

ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point(color = "steelblue") +
  geom_smooth(method = "lm", se = TRUE, color = "black") +
  labs(x = "Weight (1000 lbs)", y = "Miles per gallon") +
  theme_minimal()
```

The model explains
$R^2 = `r round(summary(model)$r.squared, 2)`$
of the variation in fuel efficiency — a strong relationship for a
single-variable model.

# Conclusion

This short analysis shows that vehicle weight is a strong predictor of
fuel efficiency: heavier cars consistently get fewer miles per gallon.
A natural next step would be to extend the model in @sec-extensions.

## Possible Extensions {#sec-extensions}

<!-- {#sec-extensions} names this heading so @sec-extensions can link here. -->

Future work could include:

- Adding horsepower and cylinder count as additional predictors
- Comparing automatic vs. manual transmissions
- Checking whether the relationship is linear or curved at extreme weights

# References {.unnumbered}

<!-- {.unnumbered} keeps this heading out of 1, 2, 3, ... even with
     number-sections: true. -->

Henderson, H. V. and Velleman, P. F. (1981). *Building multiple regression
models interactively.* Biometrics, 37, 391-411. (Source of the `mtcars`
dataset, distributed with base R.)
