---
title: "5.1 dplyr: rowwise() and across()"
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 6,
  fig.height = 3.6,
  fig.align = "center",
  dpi = 120,
  error = TRUE,
  fig.path = "_figs/dplyr-rowwise-across/"
)
library(tidyverse)
```

rowwise() summarizes across columns in one row. across() repeats a function on many columns.

## 1. `mutate()` works on whole columns

Before `rowwise()`, look at what `mutate()` normally does.

```{r}
#| label: exams
exams <- tibble(
  student = c("A", "B", "C"),
  exam1 = c(80, 70, 90),
  exam2 = c(90, 85, 95),
  exam3 = c(70, 75, 85)
)

exams |> mutate(difference = exam1 - exam2, total = exam1 + exam2 + exam3)
```

It looks as if `mutate()` subtracts row by row. It does not. Inside `mutate()`, `exam1` and `exam2` are **whole columns**, so R computes

```{r}
#| label: vector-minus
c(80, 70, 90) - c(90, 85, 95)
```

`-` and `+` are **vectorized**: they already work element by element. No `rowwise()` is needed.

---

## 2. Why `mean()` needs `rowwise()`

Now try the average score of each student:

```{r}
#| label: mean-wrong
exams |> mutate(avg = mean(c(exam1, exam2, exam3)))
```

Every row gets the same number. `c(exam1, exam2, exam3)` glues the three **whole columns** into one vector of nine values,

```{r}
#| label: nine-values
c(exams$exam1, exams$exam2, exams$exam3)
```

and `mean()` returns one mean of all nine, repeated on every row.

`rowwise()` changes the context: the next verbs run **once per row**.

```{r}
#| label: mean-right
exams |>
  rowwise() |>
  mutate(avg = mean(c(exam1, exam2, exam3))) |>
  ungroup()
```

Row 1 is `mean(c(80, 90, 70)) = 80`, row 2 is `mean(c(70, 85, 75)) = 76.67`, row 3 is `mean(c(90, 95, 85)) = 90`.

### `rowwise()` is a group of one

`group_by(carrier)` makes one group per carrier, each with many rows. `rowwise()` makes **each row its own group**. So:

- `group_by()`: the context is a group of rows.
- `rowwise()`: the context is one row.

`rowwise(name)` also keeps `name` as an identifier when you `summarize()`:

```{r}
#| label: chunk-1
df <- tibble(name = c("Mara", "Hadley"), x = 1:2, y = 3:4, z = 5:6)

df |> mutate(m = mean(c(x, y, z)))
# one mean for the whole table

df |>
  rowwise(name) |>
  summarize(m = mean(c(x, y, z)), .groups = "drop")
# one mean per row
```

Finish with `ungroup()` (or `.groups = "drop"`). Otherwise later verbs keep working one row at a time.

---

## 3. `c_across()`: the values of the current row

Typing `c(exam1, exam2, exam3)` gets long. Inside `rowwise()`, `c_across(exam1:exam3)` collects the values of those columns **in the current row** into a vector. For student A it is `c(80, 90, 70)`.

```{r}
#| label: c-across-exams
exams |>
  rowwise() |>
  mutate(
    avg_score = mean(c_across(exam1:exam3)),
    max_score = max(c_across(exam1:exam3)),
    min_score = min(c_across(exam1:exam3)),
    sd_score = sd(c_across(exam1:exam3))
  ) |>
  ungroup()
```

`c_across()` selects columns with the same helpers as `select()`:

```{r}
#| label: chunk-2
df <- tibble(id = 1:6, w = 10:15, x = 20:25, y = 30:35, z = 40:45)
rf <- df |> rowwise(id)

rf |> mutate(total = sum(c_across(w:z)))
rf |> mutate(total = sum(c_across(where(is.numeric))))
```

### Missing values

`na.rm = TRUE` works as usual. It drops the `NA`s in that row only.

```{r}
#| label: c-across-na
quizzes <- tibble(
  student = c("A", "B", "C"),
  quiz1 = c(80, 90, NA),
  quiz2 = c(70, NA, 85),
  quiz3 = c(90, 80, 95)
)

quizzes |>
  rowwise() |>
  mutate(quiz_avg = mean(c_across(quiz1:quiz3), na.rm = TRUE)) |>
  ungroup()
```

For B this is `mean(c(90, NA, 80), na.rm = TRUE) = 85`.

For speed on a whole frame, `rowSums()` and `rowMeans()` avoid the split-and-join:

```{r}
#| label: chunk-3
df |> mutate(total = rowSums(across(where(is.numeric))))
```

`starwars` (in dplyr) is the usual practice table. Without `rowwise()`, `max(height, mass, birth_year)` is one number for the whole column stack, repeated on every row.

---

## 4. `across()` repeats a function on columns

`across()` lives **inside** `mutate()`, `summarize()`, `filter()`, and the other data-masking verbs. It does not change the data by itself.

Two arguments:

- `.cols`: tidyselect, default `everything()`
- `.fns`: a function, a named list of functions, or a formula `~ median(.x, na.rm = TRUE)`. `.x` is the current column.

```{r}
#| label: chunk-4
starwars |>
  summarize(across(where(is.numeric), ~ median(.x, na.rm = TRUE)), .groups = "drop")

starwars |>
  group_by(species) |>
  filter(n() > 1) |>
  summarize(
    across(c(sex, gender, homeworld), ~ length(unique(.x))),
    n = n()
  )
```

Grouping variables are skipped so you do not summarize them by accident.

A named list makes several summaries per column:

```{r}
#| label: chunk-5
min_max <- list(
  min = ~ min(.x, na.rm = TRUE),
  max = ~ max(.x, na.rm = TRUE)
)
starwars |> summarize(across(where(is.numeric), min_max))
```

Order matters. If you compute `n = n()` first, `n` is numeric, so a later `across(where(is.numeric), sd)` will try to take the SD of that constant and give `NA`. Put `n = n()` last.

`across()` replaces the old `*_if()`, `*_at()`, and `*_all` helpers. `select()` and `rename()` already use tidyselect; use `rename_with()` to change names with a function.

---

## 5. `across()` vs `c_across()`

The names are close, the jobs are not.

- `across()` applies a function to **each selected column separately**.
- `c_across()` **combines** the selected values **within the current row**.

```{r}
#| label: xyz
xyz <- tibble(name = c("A", "B", "C"), x = c(10, 20, 30), y = c(20, 40, 60), z = c(30, 60, 90))
```

One mean **per column** (`mean(x)`, `mean(y)`, `mean(z)`):

```{r}
#| label: across-cols
xyz |> summarize(across(x:z, mean))

xyz |> mutate(across(x:z, ~ .x * 2))
```

One mean **per row** (`mean(c(10, 20, 30))`, `mean(c(20, 40, 60))`, ...):

```{r}
#| label: c-across-rows
xyz |>
  rowwise() |>
  mutate(row_mean = mean(c_across(x:z))) |>
  ungroup()
```

A picture to remember:

```
          x     y     z
        ----------------
Row 1    10    20    30   ->  c_across()
Row 2    20    40    60   ->  c_across()
Row 3    30    60    90   ->  c_across()
         |     |     |
         v     v     v
            across()
```

`across()` goes down the columns. `c_across()` goes along a row. The `c` is for **combine**, as in `c()`.

---

## 6. When you do not need `rowwise()`

Using several columns does not by itself call for `rowwise()`. These are all vectorized:

```{r}
#| label: no-rowwise
xyz |> mutate(total = x + y + z, difference = x - y, ratio = x / y, bigger = x > y)
```

`rowwise()` is for functions like `mean()`, `max()`, `min()`, `sd()` that take a vector and return **one** value, when you want that value for each row.

So do not ask "Am I using several columns?" Ask:

- Same function on **each column separately**? Use `across()`.
- **Combine values from several columns within each row**? Use `rowwise()` with `c_across()` (or `rowSums()` / `rowMeans()` for sums and means).

---

## 7. Practice

1. For each `starwars` character, the maximum of `height`, `mass`, and `birth_year` (`na.rm = TRUE`). Who is largest? Try it with and without `rowwise()`.
2. Median of every numeric column, by `species` and `gender`, plus `n()`, sorted by `n` descending.
