---
title: "1.2 R Packages and the Tidyverse"
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 6,
  fig.height = 3.6,
  fig.align = "center",
  dpi = 120,
  error = TRUE,
  fig.path = "_figs/r-packages-and-the-tidyverse/"
)
library(tidyverse)
```

Installing packages, loading them, the tidyverse, and the pipe.

## 1. What Is a Package?

R already contains many functions:

```{r}
#| label: chunk-1
mean()
sum()
sqrt()
seq()
data.frame()
```

These functions come with **Base R** and the standard R installation.

However, not every useful function is included in Base R.

R can be extended using **packages**.

> A **package** is a collection of functions, data sets, documentation, and other tools designed for particular tasks.

Thousands of R packages are available for statistics, visualization, machine learning, data manipulation, and many other applications.

---

## 2. Installing a Package

Before using a package for the first time, we usually need to **install** it.

For example:

```{r}
#| label: chunk-2
#| eval: false
install.packages("ggplot2")
```

The package name must be inside quotation marks:

```{r}
#| label: chunk-3
#| eval: false
install.packages("ggplot2")
```

not:

```{r}
#| label: chunk-4
#| eval: false
install.packages(ggplot2)
```

Installation downloads the package and places it on your computer.

### Usually, you install a package only once.

Think of installing an R package like installing an app on your computer.

---

## 3. Loading a Package

Installing a package does **not** automatically make its functions available in the current R session.

We load an installed package using:

```{r}
#| label: chunk-5
library(ggplot2)
```

Notice the difference:

```{r}
#| label: chunk-6
#| eval: false
install.packages("ggplot2")   # Install
library(ggplot2)              # Load
```

A useful rule to remember is:

> **Install once; load each new R session.**

For example, you may install `ggplot2` once:

```{r}
#| label: chunk-7
#| eval: false
install.packages("ggplot2")
```

Then, whenever you start a new R session and want to use it:

```{r}
#| label: chunk-8
library(ggplot2)
```

---

## 4. What Is an R Session?

An **R session** is the period during which R is currently running.

When you close and restart R or RStudio, you begin a new session.

Packages that you loaded previously are generally not automatically loaded into the new session.

Therefore, your script might begin with:

```{r}
#| label: chunk-9
library(ggplot2)
library(dplyr)
```

but you should **not normally put** this in a script that runs every time:

```{r}
#| label: chunk-10
#| eval: false
install.packages("ggplot2")
```

Installation is something you do when needed, not every time the script runs.

---

## 5. Using a Function from a Package

Suppose we load `ggplot2`:

```{r}
#| label: chunk-11
library(ggplot2)
```

We can now use functions provided by that package:

```{r}
#| label: chunk-12
ggplot()
```

Another way to use a function is:

```{r}
#| label: chunk-13
ggplot2::ggplot()
```

The notation:

```{r}
#| label: chunk-14
package::function()
```

means:

> "Use this function specifically from this package."

For example:

```{r}
#| label: chunk-15
dplyr::filter()
```

means:

> Use the `filter()` function from the `dplyr` package.

This can be useful because different packages can sometimes contain functions with the same name.

---

## 6. Getting Help

R has a built-in help system.

To read the documentation for a function:

```{r}
#| label: chunk-16
?mean
```

or:

```{r}
#| label: chunk-17
help(mean)
```

For a function from a package:

```{r}
#| label: chunk-18
?ggplot2::ggplot
```

You can also search the help system:

```{r}
#| label: chunk-19
help.search("mean")
```

A shortcut is:

```{r}
#| label: chunk-20
??mean
```

Therefore:

?function      → documentation for a function
??topic        → search R help for a topic


7. What Packages Are Currently Loaded?
You can see information about the current R session using:

```{r}
#| label: chunk-21
sessionInfo()
```

You can also see the current search path:

```{r}
#| label: chunk-22
search()
```

For example, after:

```{r}
#| label: chunk-23
library(ggplot2)
```

try:

```{r}
#| label: chunk-24
search()
```

and notice that `ggplot2` appears in the search path.

---

## 8. What Packages Are Installed?

To see installed packages:

```{r}
#| label: chunk-25
installed.packages()
```

This returns considerable information.

To see just their names:

```{r}
#| label: chunk-26
rownames(installed.packages())
```

To check whether a particular package is available:

```{r}
#| label: chunk-27
requireNamespace("ggplot2", quietly = TRUE)
```

---

## 9. What Is the Tidyverse?

In this course, we will learn both **Base R** and the **tidyverse**.

The tidyverse is a collection of R packages designed to work together for data science.

We can install it using:

```{r}
#| label: chunk-28
#| eval: false
install.packages("tidyverse")
```

and load it using:

```{r}
#| label: chunk-29
library(tidyverse)
```

Loading `tidyverse` loads several important packages.

Some of the most commonly used are:

Package
Main purpose




`ggplot2`
Data visualization


`dplyr`
Data manipulation


`tidyr`
Reshaping and tidying data


`readr`
Reading data files


`tibble`
Modern data frames


`stringr`
Working with strings


`forcats`
Working with categorical data/factors


`purrr`
Functional programming and iteration

We will encounter these packages gradually throughout the course.

---

## 10. Base R vs. Tidyverse

Often, the same task can be performed using either Base R or tidyverse tools.

For example, suppose:

```{r}
#| label: chunk-30
students <- data.frame(
  name = c("John", "Sara", "David"),
  grade = c(75, 92, 85)
)
```

Using **Base R**, we could select students with grades above 80:

```{r}
#| label: chunk-31
students[students$grade > 80, ]
```

or, with `subset()`:

```{r}
#| label: chunk-32
subset(students, grade > 80)
```

`subset()` lets you write the condition with the column name. In dplyr, two functions sit next to it:

- `filter()` is the closest match for keeping rows
- `select()` is the closest match for keeping columns (`subset(students, select = name)`)

```{r}
#| label: chunk-33
library(dplyr)

filter(students, grade > 80)
```

Both the Base R and dplyr approaches are valid.

Throughout this course, we will learn how to work with data using both approaches.

---

## 11. The Pipe

One important idea in modern R and the tidyverse is the **pipe**.

Suppose:

```{r}
#| label: chunk-34
x <- c(10, 20, 30, 40)
```

We could calculate its mean using:

```{r}
#| label: chunk-35
mean(x)
```

Base R provides the pipe:

```{r}
#| label: chunk-36
x |> mean()
```

You can read this approximately as:

> Take `x`, **then** calculate its mean.

The tidyverse historically popularized another pipe:

```{r}
#| label: chunk-37
x %>% mean()
```

The `%>%` pipe is provided by `magrittr` and is commonly encountered in tidyverse code.

In this course, you may encounter both:

```{r}
#| label: chunk-38
#| eval: false
|>      # Base R pipe

%>%     # traditional tidyverse pipe
```

The pipe becomes especially useful when several operations are performed in sequence.

For example:

```{r}
#| label: chunk-39
students |>
  subset(grade > 80) |>
  head()
```

`subset()` is Base R. The dplyr line that does the same row-keeping is `filter(grade > 80)`. Note 2.1 goes through `subset()`, `filter()`, and `select()` more carefully.

Later, we will use pipes extensively with tidyverse functions.

---

## 12. A Typical R Workflow

A simple R project might follow this pattern:

### Step 1 — Install required packages

Usually done only once:

```{r}
#| label: chunk-40
#| eval: false
install.packages("tidyverse")
```

### Step 2 — Load packages

Done when beginning a new R session:

```{r}
#| label: chunk-41
library(tidyverse)
```

### Step 3 — Import or create data

We can create a small data set in code:

```{r}
#| label: chunk-42
students <- data.frame(
  name = c("John", "Sara", "David"),
  grade = c(75, 92, 85)
)
```

We can also read a file. Download
[students.csv](files/data-612/students.csv)
and place it in your working directory. Then:

```{r}
#| label: chunk-43
#| eval: false
students <- read.csv("students.csv")
```

With the tidyverse, an equivalent approach is:

```{r}
#| label: chunk-44
#| eval: false
students <- readr::read_csv("students.csv")
```

`read.csv()` is Base R. `read_csv()` comes from the `readr` package.

Note **1.3** is the full treatment: relative paths, delimiters, `read.table()` / `read_csv()`, inspecting the import, and exporting.

### Step 4 — Explore the data

```{r}
#| label: chunk-45
#| eval: false
head(students)
str(students)
summary(students)
View(students)
```

### Step 5 — Manipulate the data

```{r}
#| label: chunk-46
students |>
  dplyr::filter(grade > 80)
```

### Step 6 — Analyze or visualize the data

For example:

```{r}
#| label: chunk-47
mean(students$grade)
```

or later:

```{r}
#| label: chunk-48
ggplot(students, aes(x = name, y = grade)) +
  geom_col()
```

---

## 13. Important Package Commands to Remember

```{r}
#| label: chunk-49
#| eval: false
install.packages("package")
```

**Install** a package.

```{r}
#| label: chunk-50
library(package)
```

**Load** an installed package.

```{r}
#| label: chunk-51
package::function()
```

Use a particular **function from a particular package**.

```{r}
#| label: chunk-52
?function
```

Read a function's **documentation**.

```{r}
#| label: chunk-53
??topic
```

**Search** R's help system.

```{r}
#| label: chunk-54
installed.packages()
```

See the packages that are **installed**.

```{r}
#| label: chunk-55
sessionInfo()
```

See information about the current **R session**.

```{r}
#| label: chunk-56
search()
```

See packages currently attached to the **search path**.

---

## 14. The Big Picture

A useful way to think about the R ecosystem is:

R
                         │
             ┌───────────┴───────────┐
             │                       │
           Base R                 Packages
             │                       │
     Built-in functions       Additional functions
             │                       │
     mean(), seq(), etc.      ┌──────┴─────────┐
                              │                │
                           tidyverse       Other packages
                              │
                 ┌────────────┼────────────┐
                 │            │            │
              dplyr        ggplot2       tidyr
                 │            │            │
               Data       Visualization  Reshaping
            manipulation                  data

The central distinction to remember is:

> **R is the programming language. A package extends R by providing additional functions and tools.**

And:

> **Base R and tidyverse are not two different programming languages. The tidyverse is a collection of packages that we use within R.**

For packages, remember the three most important commands:

```{r}
#| label: chunk-57
#| eval: false
install.packages("tidyverse")   # Get it
library(tidyverse)              # Load it
dplyr::filter()                 # Use a specific function
```
